An Event-based Architecture for Multi-population Optimization Algorithms

### 2020-01-19 draft.tex
Some comments and reformatting

\begin{abstract} 
Splitting a population into multiple instances is a technique that has been used extensively in
recent years to improve the performance of nature-inspired optimization
algorithms. Work on several populations can be done in parallel, and
also asynchronously, which can be leveraged to create scalable
implementations based on, among other methods, distributed, multi-threaded, parallel, and cloud-based computing. 
% I have rephrased all this. Please check it out - JJ
The design of these cloud-native, distributed, multi-population
algorithms is not a trivial task. Coming from monolithic
(single-instance) solutions, adaptations at several levels, from the
algorithmic to the functional, must be made to leverage
the scalability, elasticity, fault-tolerance, reproducibility, and
cost-effectiveness of cloud systems while, at the same time,
conserving the intended functionality. Instead of an evolutive
approach, in this paper we propose to create from scratch a
cloud-native optimization framework that is able to include multiple
algorithms and use few parameters to work. This solution goes beyond
current state of the art via a framework that is able to support
different algorithms at the same time, work asynchronously, and also
be easily deployable to any cloud platform. We evaluate the
performance and scalability of this solution, together with the effect
other design parameters had on the 
performance, in particular, the number and the size of populations
with respect to problem size. The architecture and the implemented platform is an excellent
alternative for locally or in the cloud, thus proving that cloud-native
bioinspired algorithms perform better in their "natural" environment than other
kinds of algorithms, and setting a new baseline for scaling and
performance in the cloud.
\end{abstract}

@@

\section{State of the Art} 
\label{soa}
Hadoop 
    Comparison \cite{ferrucci2018using}
    Hadoop\cite{fazenda2012} 
    FlexGP \cite{FlexGP}, 
    elephant56 \cite{salza2016elephant56},

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{kafkEOsmall}
    \caption{Example of a parametric plot ($\sin (x), \cos(x), x$)}
    \label{fig:kafkEO}
\end{figure}

@@

The primary data element used throughout the system is the subpopulation, and at a higher
level, we can view the system as a continuous stream of subpopulations flowing
from one process to the other. By "continuous stream," we mean that ideally,
everything must happen asynchronously without components needing to wait for
others. In this kind of system, this streaming functionality is normally
implemented by using a message queue system. For instance, if one process needs
to communicate a subpopulation to another, it must pack the sub-population into
a message, and then push the message asynchronously to the queuing system; this
means that after the process pushes the message, it continues its execution
without waiting for a result. On the other end, the recipient subscribes to the
message queue by defining an event handler method that is going to be triggered when a
message is pushed to the message queue. The two components at each end of the queue are
Producers that push messages to the queue and Consumers that pull them. This
pattern is an essential component of a reactive architecture because it is
highly scalable; one of the reasons for this are stateless functions. We can
define stateless functions as methods that do not have side effects, meaning
they do not need to read, keep, or modify data outside of the method. If
Consumers are implemented using stateless functions, then there is no difference
netween having one or many copies of the same function pulling work from the queue all at the
same time, because no side effects are going to result from the concurrency.
Based un these general principles we define next the proposed model.

@@

\begin{enumerate}
\item In this step, the specified number of subpopulations is created acording to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.
Each subpopulation is then pushed to the Input queue, so they can be consumed 
by Stateless functions responsible for the execution of the search.

@@

This reactive architecture has the following advantages:

An important aspect of proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e. in a virtual machine, cpu,
or thread. In this case we can have just one processing node, or many nodes, each 
running a different search strategy, or using its own parameters.

Also, the reactive controller, gives designers more control over the milti-population algorithm.
In this process, designers can dynamicly change the number of populations, population parameters,
and migration details on-the-fly.

Another advantege is that algorithm designers have many options for implementing this simple architecure. 
The same basic components can be implmented as a single multithreaded program, or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels, for a multi-threaded
execution.

On the other hand, there are several caveats as well:

It is more costly to move entire subpopulations as messages than only passing certain individuals 
from a process to another. Designers must conmsider this cost when working with large
individuals, and if possible send sub-populations using several messages, use compression or 
adjust the size of subpopulations. Pool-based algorithms also suffer from this drawback, but 
web based immplementations have been working with continous optimization problems of 1000 dimensions.
But, in other cases this is not a viable approach.

>>>>

\begin{abstract} 
Splitting a population into multiple instances is a technique that has been used extensively in
recent years to improve the performance of nature-inspired optimization
algorithms. Work on several populations can be done in parallel, and
also asynchronously, which can be leveraged to create scalable
implementations based on, among other methods, distributed, multi-threaded, parallel, and cloud-based computing. 
% I have rephrased all this. Please check it out - JJ
The design of these cloud-native, distributed, multi-population
algorithms is not a trivial task. Coming from monolithic
(single-instance) solutions, adaptations at several levels, from the
algorithmic to the functional, must be made to leverage
the scalability, elasticity, fault-tolerance, reproducibility, and
cost-effectiveness of cloud systems while, at the same time,
conserving the intended functionality. Instead of an evolutive
approach, in this paper we propose to create from scratch a
cloud-native optimization framework that is able to include multiple
algorithms and use few parameters to work. This solution goes beyond
current state of the art via a framework that is able to support
different algorithms at the same time, work asynchronously, and also
be easily deployable to any cloud platform. We evaluate the
performance and scalability of this solution, together with the effect
other design parameters had on the 
performance, in particular, the number and the size of populations
with respect to problem size. The architecture and the implemented platform is an excellent
alternative for locally or in the cloud, thus proving that cloud-native
bioinspired algorithms perform better in their "natural" environment than other
kinds of algorithms, and setting a new baseline for scaling and
performance in the cloud.
% Still a big longish. We need another iteration to reduce it - JJ
\end{abstract}

@@

\section{State of the Art} 
\label{soa}
% a lot of this has been done in the previous method... Maybe there's
% not that much to do - JJ
Hadoop 
    Comparison \cite{ferrucci2018using}
    Hadoop\cite{fazenda2012} 
    FlexGP \cite{FlexGP}, 
    elephant56 \cite{salza2016elephant56},

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{kafkEOsmall}
    \caption{Example of a parametric plot ($\sin (x), \cos(x), x$)} % This caption? - JJ
    \label{fig:kafkEO}
\end{figure}

@@

The primary data element used throughout the system is the subpopulation, and at a higher
level, we can view the system as a continuous stream of subpopulations
flowing
% We talk about "multi-population" elsewhere, including the
% abstract. We should unify terminology - JJ
from one process to the other. By "continuous stream," we mean that ideally,
everything must happen asynchronously without components needing to wait for
others. In this kind of system, this streaming functionality is normally
implemented by using a message queue system. For instance, if one process needs
to communicate a subpopulation to another, it must pack the sub-population into
a message, and then send the message asynchronously % I don't know if
                                % asynchronously applies here. It's
                                % simply sent - JJ
to the queuing system; this
means that after the process pushes the message, it continues its execution
without waiting for a result. On the other end, the recipient subscribes to the
message queue by defining an event handler method that is going to be triggered when a
message is pushed to the message queue. The two components at each end of the queue are
Producers that push messages to the queue and Consumers that pull them. This
pattern is an essential component of a reactive architecture because it is
highly scalable; one of the reasons for this are stateless functions. We can
define stateless functions as methods that do not have side effects, meaning
they do not need to read, keep, or modify data outside of the method. If
Consumers are implemented using stateless functions, then there is no difference
between having one or many copies of the same function pulling work from the queue all at the
same time, because no side effects are going to result from the
concurrency.
% This is important and we should probably mention that in the
% introduction - JJ 
Based on these general principles we define next the proposed model.

@@

\begin{enumerate}
\item In this step, the specified number of subpopulations is created acording to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.
Each subpopulation is then pushed to the Input queue, so they can be consumed 
by Stateless functions responsible for the execution of the search.  % Why the caps in Stateless? - JJ

@@

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e. in a virtual machine, cpu,
or thread. In this case we can have just one processing node, or many nodes, each 
running a different search strategy, or using its own parameters. 
\item 
Also, the reactive controller, gives designers more control over the milti-population algorithm.
In this process, designers can dynamicly change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantege is that algorithm designers have many options for implementing this simple architecure. 
The same basic components can be implmented as a single multithreaded program, or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels, for a multi-threaded
execution.
\end{itemize}

On the other hand, there are several caveats as well:
It is more costly to move entire subpopulations as messages than only passing certain individuals 
from a process to another. Designers must conmsider this cost when working with large
individuals, and if possible send sub-populations using several messages, use compression or 
adjust the size of subpopulations. Pool-based algorithms also suffer from this drawback, but 
web based immplementations have been working with continous optimization problems of 1000 dimensions.
But, in other cases this is not a viable approach.
